Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-12-03 responseToReviews.md

# Comments to the Author
## General Comments

**Wade and Baker have written a very approachable and well-motivated review of the use of VEP measurements of visual contrast. The authors rightly focus on the importance of measuring contrast responses under a wide range of conditions, highlighting the ways in which human evoked responses can be made in informative in diverse contexts. The review should thus be of broad interest.**

**There are points where the review is overly broad and others that could use more detail. These are described below.**

**A central part of the review is the role of computational models and normalization models, in particular. The use of the normalization framework for describing the VEP contrast response function (CRF) has been fruitful in terms of mechanistic interpretation and linking to other measurements, such as single-unit physiology and fMRI.  This modeling approach assumes that the CRF reflects the activity of a more or less unitary population, and that the population response resembles that of the underlying individual neurons that make up the population.**

**There are two edge-cases in the literature that suggest a more nuanced approach will be needed. It would be useful to be clearer on the assumptions of the modeling approach and what the significance is for interpretation of cases where the assumptions don’t hold.**

**Two-limbed contrast response functions. Campbell and Maffei (J. Physiol,1970) in the earliest measurements of the VEP contrast response function reported that it sometimes had two “limbs” – one at low contrast and a second one with a steeper slope at higher contrasts.  Two-limbed contrast response functions have also been found by Nakayama and Mackeben (Vision Res,1982) in macaque with electrodes on the dura, by Apkarian and Tyler (Vision Res, 1985),  Norcia et al., (Vision Res,1989) and Souza et al., (IOVS, 2007), to name a few.  As noted, the sigmoidal model assumes that the population response is unitary, but the two-limbed functions suggest that it is a mixture of populations. The two limbs may not be apparent when the function is sampled with only a few points over a wide range of contrast.**

**Harmonics with different CRFs.  Bobak et al., (IOVS, 1986), found that the first harmonic (1f) and second harmonic (2f) of the pattern onset/offset SSVEP had different thresholds but similar slopes once above threshold. Norcia et al., (TVST, 2024) found different thresholds and slope for 1f and 2f responses driven by contrast modulated dynamic noise. This again suggests the presence of multiple underlying populations.  The behavior also raises the question of whether to pool harmonics when estimating the response function. This was alluded to in the simulation, but could be expanded on.**

**The lack of dynamics in the normalization model, as presented should be discussed, the brief mention of Tsai et al. aside. The authors may wish to cite another framework that explicitly models dynamics (e.g. amplitude and phase) in the context of the contrast response function (Zemon and Gordon, Vision Res, 2006)**

### Response
We thank the reviewer for these positive comments. In general this was intended to be an overview in the sense of a 'primer' for researchers coming to the SSVEP for the first time rather than a detailed historical account of the field - hence the inclusion of code examples and relatively domain-agnostic language. However, the reviewer raises some important general issues and we are happy to address them: Specifically, we include references to the idea that SSVEP can engage more than one type of neuron, that the population responses is not trivially derived from what is known about the properties of individual neurons and that fitting with a sigmoidal response function may mask some complexities in the CRF that are more readily seen with denser sampling. We think the Bobak reference is IOVS 1984 and we have also included the Kaestner TVST 2024 reference (thank you for drawing our attention to both of these).

### Specific comments

@@

**Adaptation effects.  Adapted contrast response functions have been measured by Nelson et al., (EEG Journal, 1984) by comparing CRFs measured by sweeping contrast from low to high vs high to low. They found that responses at low contrast were smaller in the “down” sweeps and attributed it to differences in prior adaptation**

We looked hard but could not find this reference. We did find an excellent (and unexpectedly humorous) review of SSVEPs by Nelson (Nelson, J. I., Seiple, W. H., Kupersmith, M. J., & Carr, R. E. (1984). Lock-in Techniques for the Swept Stimulus Evoked Potential. Journal of Clinical Neurophysiology, 1(4), 409–436. doi:10.1097/00004691-198401040-00002) but could not find a mention of adaptation effects.

@@

Done

**2. The following statement is unclear and inaccurate: “On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on.” The term “on-off flicker” is typically used for simple luminance flicker without spatial pattern contrast. In the context of pattern stimulation, “on-off” can refer to “appearance-disappearance” of a spatial pattern with its contrast modulated in time. When the pattern appears and disappears into a uniform field of equal space-average luminance, the light and dark elements are modulated in counterphase. Any linear contribution to the response elicited by each set of elements (e.g., bars or checks) is expected to be canceled when summed at the electrode site on the scalp. Thus, the response at 1F to this stimulus is in fact a nonlinear response.**

Thank you for this explanation. We have now reworded the paper to make this clearer. We did take care to define what we mean by 'on-off' flicker in the preceding sentence but we now clarify the alternative 'appearence/disappearence' terminology. The point we were making was that odd harmonic responses (1F) are a feature of this type of modulation because the responses to the appearance and disappearance of a pattern (or mean field) need not be completely symmetrical (see Zemon, et al 1988 for an explicit demonstration).

**This type of stimulation does not separate contributions from “independent populations of on- and off-cells.”   There are other ssVEP techniques developed that do tap ON and OFF pathways selectively, which are not cited here. Contrast-reversal stimuli which involve symmetrical modulation of light and dark elements cancel odd harmonic frequency components in the response, however, both kinds of stimulus techniques mentioned here elicit only nonlinear responses.**

Apologies that we have not been sufficiently clear here. We do not think we suggested any form of 'separation' and we agree that a) there are infrequently-used paradigms that can target ON and OFF pathways more effectively (for example, 'sawtooth' temporal modulations that provide asymmetrical onset or offset transients) and b) responses from spatial features of differing polarity will tend to be combined at the scalp as long as the spatial frequency is high enough. We have expanded this section to attempt to make the general point clear.

**3. The statement “SSVEPs have proven to be an excellent measure of early chromatic processing as well” again fails to cite seminal work in this area.**

We have included three earlier papers by Regan on this topic.

**4. The statement “This normalization, achieved through a computation called ‘contrast gain control’” again neglects to cite the original work that defined “contrast gain control” and the authors ignored a nonlinear biophysical model proposed to explain this phenomenon in ssVEP contrast response functions (which was also applied in developmental work on infants and children).**

Contrast gain control is related to 'automatic gain control' - a electrical engineering concept that arose in the early 20th century. The concept of modulating the input (rather than the output) of the system became established in the 1930s and 40s (for example in WW2 radar systems). The specific idea of computing that modulation as a function of the local average signal power seems to have arisen around this time and was refined over the 50s and 60s (for example, in 'sidechain compression'). In the domain of vision, an early example of 'contrast gain control' (scaling the input - in this case luminance) is found in Sperling and Sondhi 1968.

The first formalization of the concept in the domain of spatial vision that included the critical concepts of an early, divisive gain control signal computed from some additional estimate of local contrast seems to appear in a seminal paper by Shapley and Victor, (1978). Application to contrast adaptation was demonstrated by Ohzawa, Sclar, Freeman, (1982, 1985) with a demonstration of instantaneous gain control in the human SSVEP by Bobak et al in 1988. A synthesis that included both adaptation and masking was proposed by Heeger 1992, while Foley 1994 showed an application to human psychophysical data. We have now cited these more foundational papers. We would be happy to include the relevant references to the non-linear biophysical model used in adult and developmental work if the reviewer can point us to the appropriate literature.

**5. The description of Figure 3 and the nonlinear operations performed are unclear. They state in the text that Equation 2 (“hyperbolic ratio function”) is used as a nonlinear transducer, but the figure legend states that a squaring operator was used for frequency doubling. A squaring operation would only produce a sinusoid of double the frequency of the input. The waveform shown in the left panel appears to be a full-wave rectified signal, which would contain theoretically an infinite number of higher harmonics. The text also states that “This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity.” When was full-wave rectification performed?**

We agree that we went too fast here and the description of the signal processing was confusing. We now break out the two stages into (Figure 3) full wave rectification only (to model the contributions of on and off cells) - explicitly pointing out that this produces only even harmonics but does not alter the linearity of the CRF. We then show (Figure 4) the effect of passing this full wave rectified signal through a hyperbolic ratio function and illustrate the resulting saturation of the CRF (Figure 4). We think that breaking the model out in this way makes the contribution of each part clearer and allows the reader to play with the code for separate stages. The input contrast is squared in Figure 4 simply because the exponent we choose in the saturating nonlinearity is '2' but this is a parameter that is often fitted rather than fixed and the reader is encouraged to experiment with the exponent to examine its effect on the resulting amplitude spectrum.

While there are, technically, infinite even harmonics in both Figure 3 and Figure 4, they decrease with frequency and we plot only three for reasons of space.

**6. The authors use the term “frequency tagging” which has been used in the literature to refer to stimulating different regions of the stimulus field or fellow eyes with different sinusoidal frequencies and measuring the intermodulation (sum and difference) frequency responses that result from nonlinear interactions. They do not, however, cite the foundational work in this field.**

As far as we are aware, the term refers to stimulating some *aspect* of the visual scene (a position or a feature) - not necessarily just a spatial region. There need not be more than one region (it is interesting, for example, to frequency tag a single feature or region and measure the effect of attending towards or away from it) and frequency tagged components can exist in the same location (for example, see the work on contrast normalization described above where the 'probe' and 'mask' have different tags and we explicitly examine the intermodulation terms). In fact much of this paper is concerned with discussing precisely these intermodulation terms and their origins. However, intermodulation is also not a 'necessary' part of a frequency tagging experiment: beautiful early work by Morgan et al (PNAS 1996) uses frequency tagging over space to examine attentional modulation purely in terms of the amplification or suppression of the F1 and F2 components. The tag need not even be visible - there is a fashion at the moment for 'rapid invisible frequency tagging' (RIFT) using flicker close to or beyond the CFF limit. In this sense, steady state paradigms are almost by definition frequency tagging paradigms because at least one feature or position is 'tagged'. In terms of using multiple input frequencies to characterise interactions between visual features, we think that Regan and Regan may have had the most comprehensive and some of the earliest descriptions of the method and we have credited them extensively throughout the paper. We are also happy to consider additional works if the reviewer can suggest them.

**7. There is a large body of work that used ssVEPs to investigate contrast processing developmentally (infants through adulthood) and in various disease states, which is mostly missed in this review article.**

We touch on both these areas but including them in detail would require us to extend what is already becoming a very long paper.

**8. Figures should be labeled.**

It is possible that our text compilation stage failed to auto-generate the figure labels - we apologise if this is the case. The current version seems to be working correctly (in that each Figure has a label saying 'Figure 1', 'Figure 2' etc which is referenced in the main text).

**Thus, the stated intention of this manuscript (“Here we describe how SSVEPs have been used to study visual contrast over the past 70 years”) fails to be met. Only select articles from a few sources are covered, and therefore, the richness and value of this technique is lost.**

The original document had well over 100 references and the current bibliography now contains 156 entries. Although we have tried to be fair, we have of necessity 'selected' from the literature and have doubtless omitted many important foundational works. We are happy to take suggestions if parts of the field have been misrepresented.

The original purpose of the document (now long eclipsed) was to serve as a convenient primer on the history of the technique and an interactive visualisation of key concepts implemented in Python and R using a novel 'executable paper' format. We think that getting this balance right means that we cannot include a comprehensive history of the entire field - in effect we must continue to be 'selective' and we hope that the relatively few papers we have cited here will serve as useful signposts to the richer literature.

### 2025-01-30 review.qmd
Forcing .docx to make figures. But figs not rendering properly in .pdf on website! Fix...

---
title: "Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  docx: 
    omitFigures: true
  pdf: default
    #keep-tex: true

@@

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region.](){#fig-centresurround}
:::
::: {.content-visible unless-format="docx"}
![The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region.](Figures/1_centresurround.pdf){#fig-centresurround}
:::

@@

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](){#fig-linear}
:::
::: {.content-visible unless-format="docx"}
![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](Figures/2_linearsystem.pdf){#fig-linear}
:::

@@

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE
![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](){#fig-fullwaverectification}
:::
::: {.content-visible unless-format="docx"}
![Illustration of a full-wave rectified sinusoid (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right).](Figures/3_fullwaverectification.pdf){#fig-fullwaverectification}

@@

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](){#fig-transducer}
:::
::: {.content-visible unless-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](Figures/4_transducer.pdf){#fig-transducer}
:::

@@

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](){#fig-twoinputs}
:::
::: {.content-visible unless-format="docx"}
![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](Figures/5_twoinputs.pdf){#fig-twoinputs}
:::

@@

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the linear regression fit to the lowest four data points prior to the satuating part of the curve, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](){#fig-sweepvep}
:::
::: {.content-visible unless-format="docx"}
![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the linear regression fit to the  lowest four data points prior to the satuating part of the curve, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](Figures/6_sweepplot.pdf){#fig-sweepvep}
:::

@@

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Example contrast response functions for different clinical conditions. Panel (a) shows modelled contrast response functions for epilepsy patients (blue) and control participants (black), based on the data of Tsai et al. (2011). Panel (b) shows functions for the amblyopic (red) and fellow (black) eyes of adults with amblyopia, based on the data of Baker et al. (2015). Panel (c) shows functions for children with (green) and without (black) a diagnosis of autism, based on the data of Vilidaite et al. (2018). Panel (d) shows data from *Drosophila melanogaster* (fruit flies) from Afsari et al. (2014). One day-old flies expressing a human gene linked to Parkinson's (hLRRK-G2019S) show increased SSVEP response amplitude and sensitivity (red) compared to control animals (black).](){#fig-plotclinical}
:::

>>>>

---
title: "Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  docx: 
    omitFigures: false
  pdf: default
    #keep-tex: true

@@

::: {.content-visible when-format="docx"}
![The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region.](Figures/1_centresurround.pdf){#fig-centresurround}
:::
::: {.content-visible unless-format="docx"}
![The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region.](Figures/1_centresurround.pdf){#fig-centresurround}
:::

@@

::: {.content-visible when-format="docx"}
![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](Figures/2_linearsystem.pdf){#fig-linear}
:::
::: {.content-visible unless-format="docx"}
![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](Figures/2_linearsystem.pdf){#fig-linear}
:::

@@

::: {.content-visible when-format="docx"}
![Illustration of a full-wave rectified sinusoid (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right).](Figures/3_fullwaverectification.pdf){#fig-fullwaverectification}
:::
::: {.content-visible unless-format="docx"}
![Illustration of a full-wave rectified sinusoid (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right).](Figures/3_fullwaverectification.pdf){#fig-fullwaverectification}

@@

::: {.content-visible when-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](Figures/4_transducer.pdf){#fig-transducer}
:::
::: {.content-visible unless-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](Figures/4_transducer.pdf){#fig-transducer}
:::

@@

::: {.content-visible when-format="docx"}
![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](Figures/5_twoinputs.pdf){#fig-twoinputs}
:::
::: {.content-visible unless-format="docx"}
![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](Figures/5_twoinputs.pdf){#fig-twoinputs}
:::

@@

::: {.content-visible when-format="docx"}
![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the linear regression fit to the  lowest four data points prior to the satuating part of the curve, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](Figures/6_sweepplot.pdf){#fig-sweepvep}
:::
::: {.content-visible unless-format="docx"}
![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the linear regression fit to the  lowest four data points prior to the satuating part of the curve, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](Figures/6_sweepplot.pdf){#fig-sweepvep}
:::

@@

::: {.content-visible when-format="docx"}
![Example contrast response functions for different clinical conditions. Panel (a) shows modelled contrast response functions for epilepsy patients (blue) and control participants (black), based on the data of Tsai et al. (2011). Panel (b) shows functions for the amblyopic (red) and fellow (black) eyes of adults with amblyopia, based on the data of Baker et al. (2015). Panel (c) shows functions for children with (green) and without (black) a diagnosis of autism, based on the data of Vilidaite et al. (2018). Panel (d) shows data from *Drosophila melanogaster* (fruit flies) from Afsari et al. (2014). One day-old flies expressing a human gene linked to Parkinson's (hLRRK-G2019S) show increased SSVEP response amplitude and sensitivity (red) compared to control animals (black).](Figures/7_clinicalplot.pdf){#fig-plotclinical}
:::
